---
title: "Bayes for BIOSTATS690P: Getting set up"
output:
  pdf_document: default
  html_notebook: default
---

Goals: 

- install R package brms and a C++ compiler
- test that your installation is working 



# Installation (do this once)
Follow the instructions from the brms readme: https://github.com/paul-buerkner/brms#faq

1. Install the R package brms, i.e. 
```{r}
#install.packages("brms", lib = "C:/Program Files/R/R-4.0.5/library")

#.libPaths()

myPaths <- .libPaths()   # get the paths
myPaths <- c(myPaths[2], myPaths[1])  # switch them
.libPaths(myPaths)  # reassign them

```

2. Install a C++ compiler:
- Windows: The program Rtools (available on https://cran.r-project.org/bin/windows/Rtools/) comes with a C++ compiler for Windows. 
- Mac: Install Xcode. 

For further instructions 
on how to get the compilers running, see the prerequisites section on
https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started.

```{r}
remove.packages("rstan", lib = "C:/Program Files/R/R-4.0.5/library")
remove.packages("rstan")
if (file.exists(".RData")) file.remove(".RData")
```

```{r}
Sys.setenv(DOWNLOAD_STATIC_LIBV8 = 1) # only necessary for Linux without the nodejs library / headers
install.packages("rstan", repos = "https://cloud.r-project.org/", dependencies = TRUE)
```

```{r}
example(stan_model, package = "rstan", run.dontrun = TRUE)
```

```{r}
remotes::install_github("paul-buerkner/brms")
```

```{r}
(myPaths <- .libPaths())   # get the paths
myPaths <- c(myPaths[2], myPaths[1])  # switch them
.libPaths(myPaths)  # reassign them
```




```{r}
#Compile packages using all cores
#Sys.setenv(MAKEFLAGS = paste0("-j",parallel::detectCores()))

#install.packages(c("StanHeaders","rstan"),type="source")
```



```{r}
example(stan_model, package = "rstan", run.dontrun = TRUE)
```





```{r}
Sys.which("make")

```


```{r}
library(brms)
library(rstan, lib.loc = "C:/Program Files/R/R-4.0.5/library")


# Compile packages using all cores
Sys.setenv(MAKEFLAGS = paste0("-j",parallel::detectCores()))

install.packages(c("StanHeaders","rstan"),type="source", lib.loc = "C:/Program Files/R/R-4.0.5/library")
```



# Test that your installation is ready to go

Load the library
```{r}
library(brms)
```

Using the example from the brms readme: 

Fit a model (takes a little while)
```{r}
library(brms)
#install.packages("plyr", lib = "C:/Program Files/R/R-4.0.5/library")
fit1 <- brm(count ~ zAge + zBase * Trt + (1|patient), 
            data = epilepsy, family = poisson())
```

and check some output
```{r}
summary(fit1) 
plot(fit1, pars = c("Trt", "zBase")) 
plot(conditional_effects(fit1, effects = "zBase:Trt"))
```

The end! 
