---
title: "Review of regression in R"
output:
  html_document:
    df_print: paged
---


```{r, message = FALSE}
library(tidyverse)
# two packages that make it easier to work with output from regression models
library(pander)
library(broom)
```

# HERS data
The data used here were collected from the heart and estrogen/progestin study (HERS), in which a clinical trial of hormone therapy (HT) was conducted for prevention of recurrent heart attacks and death among 2,763 post-menopausal women with existing coronary heart
disease (CHD). 

Reading in the data, and removing missing values: 
```{r}
hers_na <- read.csv("data/hers.csv", header = TRUE)
hers <- na.omit(hers_na)
```

# Some exploratory plots

SBP plotted against age
```{r, message = FALSE}
ggplot(data = hers, mapping = aes(x = age, y = SBP)) + 
  geom_point(size = 0.5, alpha = 0.5)
```

Now using jittering and adding a smoother
```{r, message = FALSE}
ggplot(data = hers, mapping = aes(x = age, y = SBP)) + 
  geom_point(size = 0.5, alpha = 0.5, position = "jitter") +
  geom_smooth(method = "loess", color = "red")  #adding a smooth line
```

# Regression model 

Example 1: simple linear regression
```{r}
mod <- lm(SBP ~ age, data = hers)
```

Easy ways to get information on the fitted model:
Using pander
```{r}
mod %>% pander
```

Or, if you want the output as a tibble:
```{r}
mod %>% tidy()
```

Showing model fit using geom_smooth option
```{r, message = FALSE}
ggplot(data = hers, mapping = aes(x = age, y = SBP)) + 
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "lm", color = "red")  #adding a regression line
```
# Example of binary outcome data using WCGS data

Read the data
```{r}
wcgs <- read.csv("data/wcgs.csv", header = TRUE) %>%
  mutate(chd = ifelse(chd69 == "No", 0, 1))
```

Plot
```{r}
ggplot(data = wcgs, mapping = aes(x = age, y = chd)) + 
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess", color = "red")  
```

```{r}
logitreg <- glm(chd ~ age, data = wcgs, family = binomial)
logitreg
#must include the family option which indicates the probability distribution
summary(logitreg) %>% pander
```